Gametic Selection, Meiotic Drive, Sex Ratio Bias, and Transitions Between Sex Determination Systems

### 2017-02-10 LaTeX/Turnover.tex
fine editing mikes final version to send to sally

It has been suggested that sex ratio selection could be a particularly important force driving transitions between sex-determining systems \citep[Chapter 7]{Beukeboom:2014vb}. 
The default mode of sex ratio selection is `Fisherian' sex ratio selection, which favours equal investment in male and female offspring \citep[i.e., a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce,][]{Fisher:1930wy,Charnov:1982wg}. 
Given that the sex determination system can directly affect the sex ratio, we might expect Fisherian sex ratio selection to influence the spread of new sex determination systems. 
For example, \citet{Kozielska:2010vm} consider systems in which the ancestral sex chromosomes experience meiotic drive (e.g., where driving X or Y chromosomes are inherited disproportionately often), which causes sex ratios to become biased \citep{Hamilton:1967ts}. 
They find that new, unlinked sex-determining loci (masculinizing or feminizing mutations, i.e., neo-Y or neo-W loci) can then spread, which restoring an even sex ratio.

We note two other ways in which sex determination has been shown to relate to zygotic sex ratios.
Firstly, female-biased sex ratios can be favoured when there is local mate competition, in which brothers compete for mates \citep{Hamilton:1967ts}. 
Therefore, with local mate competition, feminizing mutations can spread because they bias the sex ratio towards females \citep{Wilson:1981vm,Vuillleumier:2007bh}. 
Secondly, environmental conditions (e.g., maternal condition, mate quality, age, or host size) can differentially affect the fitness of males versus females such that the optimal allocation to males/females depends on the environment \citep{Trivers:1973wb,Charnov:1977tx,Charnov:1982wg}. 
In such cases, flexible sex determination systems may evolve in order to allow the zygotic sex ratio to be determined in a way that depends on the environment \citep{Charnov:1977tx,Werren:1984tl,Pen:2010kk}. 
In this study, we do not consider environmental condition dependence or local mate competition \citep[reviewed in][]{Charnov:1982wg,Bull:1983vi,West:2009we}.

%\textcolor{red}{We add haploid selection:}
Here, we use mathematical models to find the conditions under which new sex determination systems are favoured when loci experience haploid selection. 
Haploid genotypes at many loci experience selection during gamete competition and/or meiotic drive \citep{Mulcahy:1996ha,JOSEPH:2004haa}.
We use the term `meiotic drive' to refer to the biased (non-medelian) segregation of genotypes during gamete production and the term `gametic competition' to refer to selection upon haploid genotypes within a gamete/gametophyte pool; the term `haploid selection' encompasses both processes. 
Meiotic drive generally occurs either during the production of male or female gametes only \citep{Ubeda:2005gw,Lindholm:2016cw}.
Because there are typically more pollen/sperm than required for fertilization, gametic competition is also typically sex specific, occurring primarily among male gametes.
Gametic competition may be particularly common in plants, in which 60-70\% of all genes are expressed in the male gametophyte and these genes exhibit stronger signatures of selection than random genes \citep{Borg:2009jpa,Arunkumar:2013iq,Gossmann:2014dua}.
In addition, artificial selection pressures applied to male gametophytes cause the frequency of resistant alleles to increase \citep[e.g.,][]{Hormaza:1996cv,Ravikumar:2003uo,Hedhly:2004iv,Clarke:2004ir}. 
A much smaller proportion of genes are thought to be expressed and selected during competition in animal sperm, although precise estimates are uncertain \citep{Zheng:2001fi,JOSEPH:2004haa,Vibranovski:2010et}.

@@

Surprisingly, our models show that haploid selection influences the evolution of new sex determination systems in a way that is distinct from both diploid sex-specific selection and Fisherian sex ratio selection. 
We find that the spread of new sex determination systems are independent of there being a zygotic sex ratio bias caused by associations between sex-determining regions and haploid selected loci. 
In addition, we find that associations that build up between an ancestral sex-determining locus and a haploid-selected locus can favour transitions between male and female heterogamety (e.g., a neo-W allele arising at a previously autosomal locus spreads in an ancestrally XY system), despite the fact that these ancestral associations were built up by selection. 
This does not occur in models that do not include haploid selection.

@@

%life-cycle
In each generation, we census the genotype frequencies in male and female gametes/gametophytes (hereafter gametes) before gametic competition. 
A full description of our model, including recursion equations, is given in the Appendix. 
First, competition occurs among male gametes (sperm/pollen competition) and among female gametes (egg/ovule competition) separately. 
Selection during gametic competition depends on the \textbf{A} locus genotype, relative fitnesses are given by $w_A^\Hermaphrodite$ and $w_a^\Hermaphrodite$ ($\Hermaphrodite \in \{\female,\male\}$; see table \ref{tab:fitnesstable}). %$w_{A}^\male$ and $w_{a}^\male$ for male gametes and $w_{A}^\female$ and $w_{a}^\female$ for female gametes, see table \ref{tab:fitnesstable}. 
We assume that all gametes compete for fertilization during gametic competition, which is not the case for monogamous mating systems where gametes from only one mating partner are present. 
Gametic competition in monogamous mating systems is equivalent to meiotic drive in our model, which only alters the frequency of gametes produced by heterozygotes. 
After gametic competition, random mating occurs between male and female gametes.
The resulting zygotes develop as males or females, depending on their genotypes at the \textbf{X} and \textbf{M} loci (and the \textbf{M} genotype of their mother in the case of maternal control) as described above.
Diploid males and females then experience selection, relative fitnesses are given by $w_{g}^{\male}$ in males and $w_{g}^{\female}$ in females, where $g$ is the diploid genotype at the \textbf{A} locus ($g \in \{AA, Aa, aa\}$).  
%Diploids compete with others of the same sex, where selection depends on the sex of the individual and its genotype at the \textbf{A} locus.
The next generation of gametes are then produced by meiosis, during which recombination and sex-specific meiotic drive can occur. 
Recombination occurs between loci \textbf{X} and \textbf{A} with probability $r$, between loci \textbf{A} and \textbf{M} with probability $R$, and between loci \textbf{X} and \textbf{M} with probability $\chi$.
Therefore, any order of the loci can be modelled with appropriate choices of $r$, $R$, and $\chi$ (see Table \ref{tab:chisubstitutions}). 
Males/females that are heterozygous at the \textbf{A} locus experience meiotic drive; $Aa$ heterozgotes of sex $\Hermaphrodite$ produce gametes bearing allele $A$ with probability $\alpha^\Hermaphrodite$. 
Thus, the \textbf{A} locus can experience sex-specific gametic competition, diploid selection and/or meiotic drive.

@@

\subsection*{Turnover between sex-determination systems} 
The evolution of a new sex determination system requires that a rare mutant allele, $m$, at the novel sex-determining locus  increases in frequency when rare. 
The spread of a rare mutant $m$ at the \textbf{M} locus is determined by the leading eigenvalue, $\lambda$, of the system described by the next generation frequency of eggs and sperm carrying the mutation, (\ref{eq:recursions}c), (\ref{eq:recursions}d), (\ref{eq:recursions}g), (\ref{eq:recursions}h), which is an eight equation system. %, evaluated at the above equilibrium.
%Thus, the system describing the change in frequency of the $m$ allele consists of eight recursion equations. 
Dominant neo-Y (when $k=0$) or neo-W alleles (when $k=1$) are only found in male diploids (neo-Y) or female diploids (neo-W) such that their growth rate ultimately depends only on the change in frequency of $m$-bearing gametes produced by males (for a neo-Y) or by females (for a neo-W). 
Furthermore, if the $m$ allele is fully dominant over the ancestral sex-determining system, phenotypes are not affected by the genotype at the ancestral sex-determining region (\textbf{X} locus). 
Thus, the invasion of rare dominant neo-Y or neo-W alleles is determined by the largest eigenvalue that solves the quadratic characteristic polynomial
$\lambda^2+ b \lambda + c =0$.
%\begin{equation}
%\lambda^2+ b \lambda + c =0
%\label{eq:charpoly_neoY}
%\end{equation}
%
%\noindent
%where $b$ is the average of the growth rates of the two haplotypes that carry the $m$ allele ($mA$ and $ma$), $b=(\lambda_{mA}+\lambda_{ma})/2$, and $c$ also involves the fitness of $m$ alleles when they recombine onto the other \textbf{A} background in a heterozygote, $c=\lambda_{mA}\lambda_{ma}+\rho_{mA}\rho_{ma}$ (see table \ref{tab:haplotype_growth}).
In this case $b= - (\lambda_{mA} + \lambda_{ma})$ and $c = \lambda_{mA} \lambda_{ma} -\rho_{mA} \rho_{ma}$, where $\lambda_{mi}$ \textcolor{red}{I think we can call this growth rate, rather than reserving "growth rate" for $\lambda-1$. e.g., in ecological models, can't we call both R (in discrete time) and r (in continuous time) "growth rates"?} is the growth rate of mutant haplotypes on background $i\in\{A,a\}$, accounting for loss due to recombination, and $\rho_{mi}$ is the rate of addition of mutant haplotypes onto background $i\in\{A,a\}$ due to recombination (see table \ref{tab:haplotype_growth}).
\textcolor{red}{check these interpretations. the $\lambda$s are certainly related to true growth rates because we get them by looking at the case where $R=0$ so that you only have the haplotypes growing/declining. }
% as there are now two 2s in the lambdas of table 2. perhaps the 2 in the denominator is needed because the frequency of males sums to one as does the frequency of females?}

@@

Equation \eqref{eq:lambda_neoW} shows that if there is no selection upon haploid genotypes ($t^\Hermaphrodite=\alpha^\Hermaphrodite_{\Delta}=0$), as considered by \citet{vanDoorn:2010hu}, the spread of a neo-W is equivalent to the spread of a neo-Y ($\lambda_{W',XY}=\lambda_{Y',XY}$) such that heterogametic transitions (XY to ZW or ZW to XY) can also occur only if the neo-sex-determining region is more closely linked to a locus under selection ($R<r$). 
However, if there is any haploid selection, the additional term in equation \eqref{eq:lambda_neoW} can be positive, which can allow neo-W invasion ($\lambda_{W',XY}>1$) even when the neo-sex-determining region is less closely linked to the selected locus ($R>r$). 
These transitions are unusual because, when $R>r$, associations that build up by selection between sex and selected alleles will be weakened. 
Therefore, mean fitness can decrease, see Figure \ref{fig:Combination_MeanFit}B,D.

We find that neo-W alleles can invade for a large number of selective regimes. 
To clarify the parameter space under which $\lambda_{W',XY}>1$, we consider several special cases. 
Firstly, if the \textbf{A} locus is unlinked to the ancestral sex-determining region ($r=1/2$), a more closely linked neo-W ($R<1/2$) can always invade because $\left( \hat{p}^\male_Y-\hat{p}^\male_X \right)=0$ such that the second term in equation \eqref{eq:lambda_neoW} disappears and invasion depends only on the sign of $(r-R)$. 
Indeed, invasion typically occurs when the neo-W is more closely linked to the selected locus than the ancestral sex-determining region (Figure \ref{fig:Combination_Centimorgans}).
Secondly, we can simplify cases where invasion occurs despite $R>r$ using the special case where $R=1/2$ and $r<1/2$. 
In table \ref{tab:specialcases} we give the conditions where invasion occurs where we further assume that haploid selection only occurs during one phase in one sex (e.g., during male meiosis only) and dominance coefficients are equal in the two sexes, $h^\female=h^\male$. 
Where there is no gametic competition and meiotic drive in one sex only, an unlinked neo-W can invade as long as the same allele is favoured in male and female diploid selection ($s^\female s^\male>0$, see Figure \ref{fig:Combination_Centimorgans}B), which is 50\% of the parameter space. 
Where there is no meiotic drive and gametic competition occurs in one sex only, an unlinked neo-W can invade as long as the same allele is favoured in male and female diploid selection and there are sex differences in selection of one type (e.g., $s^\female(s^\male-s^\female)>0$, see Figure \ref{fig:Combination_Centimorgans}C,D), which is 25\% of the parameter space. 
These special cases indicate that neo-W invasion can occur for a relatively large fraction of parameter space, even if the neo-W is less tightly linked to the selected locus, $R>r$.

@@

Under Fisherian sex ratio selection, autosomal modifiers favour equal investment in male and female offspring, i.e., a 1:1 sex ratio \citep{Fisher:1930wy,Charnov:1982wg,West:2009we}. 
Therefore, we are particularly interested in the case where the novel environmental sex-determiner causes half of its carriers to become female and half to become male ($k=1/2$). 
As with an autosomal locus, these sex-determiners will be found in males half of the time and in females half of the time.
However, we find that the growth rate of a rare, dominant offspring-controlled neo-ESD allele that produces males or females with equal probability ($k=1/2$) is
\begin{equation}
\lambda_{ESD',XY} =1+ \frac{1}{2}\frac{(\lambda_{Y',XY}-1) + (\lambda_{W',XY}-1)}{2} \Big|_{R=1/2} + O\left(\epsilon^3\right),
%V_A \frac{r}{1-2r}  \frac{s_{f} \left( s^\male - 3 s^\female\right) }{8 \left( s_{f} + s_{m} \right)^2} t_{m}^2.
\label{eq:lambda_ESD}
\end{equation}
\noindent
We note that recombination between the selected locus and the novel sex-determining locus, $R$, doesn't enter into the $k=1/2$ results because sex is essentially randomized each generation, %as if the novel sex-determiner were on an unlinked site that half the time acts as a neo-$W$ and half the time as a neo-$Y$.
preventing associations from building up between allele $A$ and sex.  
Equation \eqref{eq:lambda_ESD} shows that invasion by a novel `perfect' ESD (equal sex ratio, $k=1/2$) mutation is the same for an ancestrally XY or ZW system (since $\lambda_{Y',XY} = \lambda_{W',ZW}$, $\lambda_{W',XY} = \lambda_{Y',ZW}$).
Thus, by the same argument as above (if drive occurs in males only then the sex ratio is only biased when the ancestral sex-determination system is XY), Fisherian sex ratio selection does not explain invasion by an offspring-controlled neo-ESD locus. 
Rather, the neo-ESD gets half of the fitness of a feminizing mutation (neo-$W$) and half of the fitness of a masculizing mutation (neo-$Y$), but only has an effect one half of the time (the other half of the time it produces the same sex as the ancestral system would have, to leading order). 
The net result can be that perfect ESD will not invade, even if sex ratios are biased. 
For example, if there is haploid selection in males (either drive or pollen/sperm competition) but the conditions in table \ref{tab:specialcases} are not met, perfect ESD will not invade, even though it would equalize the zygotic sex ratio from an initially biased case (assuming $r<1/2$).

@@

Fisherian sex ratio selection is sometimes considered in terms of balancing parental investment in male versus female offspring \citep{Charnov:1982wg}.
In addition, under environmental sex-determination, the proportion of males/females is sometimes controlled by the mother, e.g., the proportion of eggs laid in warm versus cold environments could determine the sex ratio of offspring. 
We therefore also considered the invasion of a neo-sex-determining allele ($m$) in a model in which mothers that have at least one $m$ allele produce $k$ daughters and $(1-k)$ males. 
As with offspring-controlled ESD, for all $k\in\{0,1/2,1\}$, we find that invasion into an ancestrally XY system is the same as invasion into an ancestrally ZW system (at least up to order $\epsilon^3$), implying transitions between genetic sex determination and maternally controlled environmental sex determination are not driven by Fisherian sex ratio selection on biased zygotic sex ratios.

@@

Taken at face value, our results indicate that transitions in heterogametey (XY to ZW or vice versa) are more likely to be favoured by selection if there is selection upon both haploid and diploid genotypes rather than diploid selection alone. 
This prediction could be examined using a suitable proxy for haploid selection, for example, \citep{Lenormand:2005vb} use the outcrossing rate as a proxy for the strength of pollen competition. 
In animals, one might expect gametic competition to be stronger in species where sperm is required to live for a long time after spermatogenesis because transcripts shared during spermatogenesis may become depleted, revealing the haploid phenotype of the sperm \citep{Immler:2014im}. 
Given the caveats mentioned above about the form of meiotic drive modelled, we would also expect that heterogametic transitions in sex determination would be more common in clades where there is meiotic drive.

@@

Selection among diploids then occurs according to the diploid genotype at the \textbf{A} locus, $h$, for an individual of type $ij$ ($h \in AA$, $Aa$, $aa$, see Table \ref{tab:fitnesstable}). 
The diploid frequencies after selection in sex $d$ are given by $xx_{ij}^{\Hermaphrodite,s}=w_{h}^{\Hermaphrodite} xx_{ij}/\bar{w}^{\Hermaphrodite}$, $xy_{ij}^{\Hermaphrodite,s}=w_{h}^{\Hermaphrodite} xy_{ij}/\bar{w}^{\Hermaphrodite}$, and $yy_{ij}^{\Hermaphrodite,s}=w_{h}^{\Hermaphrodite} yy_{ij}/\bar{w}^{\Hermaphrodite}$, where $\bar{w}^{\Hermaphrodite}= \sum_{i=1}^{4}\sum_{j=1}^{4}w_{h}^{\Hermaphrodite}xx_{ij}+w_{h}^{\Hermaphrodite}xy_{ij}+w_{h}^{\Hermaphrodite}yy_{ij}$ is the mean fitness of individuals of sex $d$.

>>>>

It has been suggested that sex ratio selection could be a particularly important force driving transitions between sex-determining systems \citep[Chapter 7]{Beukeboom:2014vb}. 
The default mode of sex ratio selection is `Fisherian' sex ratio selection, which favours equal investment in male and female offspring \citep[i.e., a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce,][]{Fisher:1930wy,Charnov:1982wg}. 
Given that the sex determination system can directly affect the sex ratio, we might expect Fisherian sex ratio selection to influence the spread of new sex determination systems. 
For example, \citet{Kozielska:2010vm} consider systems in which the ancestral sex chromosomes experience meiotic drive (e.g., where driving X or Y chromosomes are inherited disproportionately often), which causes sex ratios to become biased \citep{Hamilton:1967ts}. 
They find that new, unlinked sex-determining loci (masculinizing or feminizing mutations, i.e., neo-Y or neo-W loci) can then spread, which restore an even sex ratio.

We note two other ways in which sex determination has been shown to relate to zygotic sex ratios.
Firstly, female-biased sex ratios can be favoured when there is local mate competition, where all matings are between siblings and assuming one male can inseminate many females \citep{Hamilton:1967ts}. 
Therefore, with local mate competition, feminizing mutations can spread because they bias the sex ratio towards females \citep{Wilson:1981vm,Vuillleumier:2007bh}. 
Secondly, environmental conditions (e.g., maternal condition, mate quality, age, or host size) can differentially affect the fitness of males versus females such that the optimal allocation to males/females depends on the environment \citep{Trivers:1973wb,Charnov:1977tx,Charnov:1982wg}. 
In such cases, flexible sex determination systems may evolve in order to allow the zygotic sex ratio to be determined in a way that depends on the environment \citep{Charnov:1977tx,Werren:1984tl,Pen:2010kk}. 
In this study, we do not consider environmental condition dependence or local mate competition \citep[reviewed in][]{Charnov:1982wg,Bull:1983vi,West:2009we}.

%\textcolor{red}{We add haploid selection:}
Here, we use mathematical models to find the conditions under which new sex determination systems are favoured when loci experience haploid selection. 
Haploid genotypes at many loci experience selection during gamete competition and/or meiotic drive \citep{Mulcahy:1996ha,JOSEPH:2004haa}.
We use the term `meiotic drive' to refer to the biased (non-Mendelian) segregation of genotypes during gamete production and the term `gametic competition' to refer to selection upon haploid genotypes within a gamete/gametophyte pool; the term `haploid selection' encompasses both processes. 
Meiotic drive generally occurs either during the production of male or female gametes only \citep{Ubeda:2005gw,Lindholm:2016cw}.
Because there are typically many more pollen/sperm than required for fertilization, gametic competition is also typically sex specific, occurring primarily among male gametes.
Gametic competition may be particularly common in plants, in which 60-70\% of all genes are expressed in the male gametophyte and these genes exhibit stronger signatures of selection than random genes \citep{Borg:2009jpa,Arunkumar:2013iq,Gossmann:2014dua}.
In addition, artificial selection pressures applied to male gametophytes cause the frequency of resistant alleles to increase \citep[e.g.,][]{Hormaza:1996cv,Ravikumar:2003uo,Hedhly:2004iv,Clarke:2004ir}. 
A much smaller proportion of genes are thought to be expressed and selected during competition in animal sperm, although precise estimates are uncertain \citep{Zheng:2001fi,JOSEPH:2004haa,Vibranovski:2010et}.

@@

Surprisingly, our models show that haploid selection influences the evolution of new sex determination systems in a way that is distinct from both diploid sex-specific selection and Fisherian sex ratio selection. 
We find that the spread of new sex determination systems are independent of there being a zygotic sex ratio bias caused by associations between sex-determining regions and haploid selected loci. 
In addition, we find that associations that build up between an ancestral sex-determining locus and a haploid-selected locus can favour sex chromosome transitions between male and female heterogamety (e.g., a neo-W allele arising at a previously autosomal locus spreads in an ancestrally XY system), despite the fact that these ancestral associations were built up by selection. 
This does not occur in models that do not include haploid selection.

@@

%life-cycle
In each generation, we census the genotype frequencies in male and female gametes/gametophytes (hereafter gametes) before gametic competition. 
A full description of our model, including recursion equations, is given in the Appendix. 
First, competition occurs among male gametes (sperm/pollen competition) and among female gametes (egg/ovule competition) separately. 
Selection during gametic competition depends on the \textbf{A} locus genotype, relative fitnesses are given by $w_A^\Hermaphrodite$ and $w_a^\Hermaphrodite$ ($\Hermaphrodite \in \{\female,\male\}$; see table \ref{tab:fitnesstable}). %$w_{A}^\male$ and $w_{a}^\male$ for male gametes and $w_{A}^\female$ and $w_{a}^\female$ for female gametes, see table \ref{tab:fitnesstable}. 
We assume that all gametes compete for fertilization during gametic competition, which is not the case for monogamous mating systems where gametes from only one mating partner are present. 
Gametic competition in monogamous mating systems is equivalent to meiotic drive in our model, which only alters the frequency of gametes produced by heterozygotes. 
After gametic competition, random mating occurs between male and female gametes.
The resulting zygotes develop as males or females, depending on their genotypes at the \textbf{X} and \textbf{M} loci (and the \textbf{M} genotype of their mother in the case of maternal control) as described above.
Diploid males and females then experience selection, relative fitnesses are given by $w_{g}^{\male}$ in males and $w_{g}^{\female}$ in females, where $g$ is the diploid genotype at the \textbf{A} locus ($g \in \{AA, Aa, aa\}$).  
%Diploids compete with others of the same sex, where selection depends on the sex of the individual and its genotype at the \textbf{A} locus.
The next generation of gametes are then produced by meiosis, during which recombination and sex-specific meiotic drive can occur. 
Recombination (i.e., an odd number of cross-overs) occurs between loci \textbf{X} and \textbf{A} with probability $r$, between loci \textbf{A} and \textbf{M} with probability $R$, and between loci \textbf{X} and \textbf{M} with probability $\chi$.
Therefore, any order of the loci can be modelled with appropriate choices of $r$, $R$, and $\chi$ (see Table \ref{tab:chisubstitutions}). 
Males/females that are heterozygous at the \textbf{A} locus experience meiotic drive; $Aa$ heterozgotes of sex $\Hermaphrodite$ produce gametes bearing allele $A$ with probability $\alpha^\Hermaphrodite$. 
Thus, the \textbf{A} locus can experience sex-specific gametic competition, diploid selection and/or meiotic drive.

@@

\subsection*{Turnover between sex-determination systems} 
The evolution of a new sex determination system requires that a rare mutant allele, $m$, at the novel sex-determining locus  increases in frequency when rare. 
The spread of a rare mutant $m$ at the \textbf{M} locus is determined by the leading eigenvalue, $\lambda$, of the system described by the next generation frequency of eggs and sperm carrying the mutation, (\ref{eq:recursions}c), (\ref{eq:recursions}d), (\ref{eq:recursions}g), (\ref{eq:recursions}h), which is an eight equation system. %, evaluated at the above equilibrium.
%Thus, the system describing the change in frequency of the $m$ allele consists of eight recursion equations. 
Dominant neo-Y (when $k=0$) or neo-W alleles (when $k=1$) are only found in male diploids (neo-Y) or female diploids (neo-W) such that their growth rate ultimately depends only on the change in frequency of $m$-bearing gametes produced by males (for a neo-Y) or by females (for a neo-W). 
Furthermore, if the $m$ allele is fully dominant over the ancestral sex-determining system, phenotypes are not affected by the genotype at the ancestral sex-determining region (\textbf{X} locus). 
Thus, the invasion of rare dominant neo-Y or neo-W alleles is determined by the largest eigenvalue that solves the quadratic characteristic polynomial
$\lambda^2+ b \lambda + c =0$.
%\begin{equation}
%\lambda^2+ b \lambda + c =0
%\label{eq:charpoly_neoY}
%\end{equation}
%
%\noindent
%where $b$ is the average of the growth rates of the two haplotypes that carry the $m$ allele ($mA$ and $ma$), $b=(\lambda_{mA}+\lambda_{ma})/2$, and $c$ also involves the fitness of $m$ alleles when they recombine onto the other \textbf{A} background in a heterozygote, $c=\lambda_{mA}\lambda_{ma}+\rho_{mA}\rho_{ma}$ (see table \ref{tab:haplotype_growth}).
In this case $b= - (\lambda_{mA} + \lambda_{ma})$ and $c = \lambda_{mA} \lambda_{ma} -\rho_{mA} \rho_{ma}$, where $\lambda_{mi}$ is the (discrete time) growth rate of mutant haplotypes on background $i\in\{A,a\}$, accounting for loss due to recombination, and $\rho_{mi}$ is the rate of addition of mutant haplotypes onto background $i\in\{A,a\}$ due to recombination (see table \ref{tab:haplotype_growth}).
\textcolor{red}{check these interpretations (note the 2s in the lambdas of table 2). the $\lambda$s are certainly related to true growth rates because we get them by looking at the case where $R=0$ so that you only have the haplotypes growing/declining. }
% as there are now two 2s in the lambdas of table 2. perhaps the 2 in the denominator is needed because the frequency of males sums to one as does the frequency of females?}

@@

Equation \eqref{eq:lambda_neoW} shows that if there is no selection upon haploid genotypes ($t^\Hermaphrodite=\alpha^\Hermaphrodite_{\Delta}=0$), as considered by \citet{vanDoorn:2010hu}, the spread of a neo-W is equivalent to the spread of a neo-Y ($\lambda_{W',XY}=\lambda_{Y',XY}$) such that heterogametic transitions (XY to ZW or ZW to XY) can also occur only if the neo-sex-determining region is more closely linked to a locus under selection ($R<r$). 
However, if there is any haploid selection, the additional term in equation \eqref{eq:lambda_neoW} can be positive, which can allow, for example, neo-W invasion ($\lambda_{W',XY}>1$) even when the neo-sex-determining region is less closely linked to the selected locus ($R>r$). 
These transitions are unusual because, when $R>r$, associations that build up by selection between sex and selected alleles will be weakened. 
Therefore, mean fitness can decrease, see Figure \ref{fig:Combination_MeanFit}B,D.

We find that neo-W alleles can invade an XY system for a large number of selective regimes. 
To clarify the parameter space under which $\lambda_{W',XY}>1$, we consider several special cases. 
Firstly, if the \textbf{A} locus is unlinked to the ancestral sex-determining region ($r=1/2$), a more closely linked neo-W ($R<1/2$) can always invade because $\left( \hat{p}^\male_Y-\hat{p}^\male_X \right)=0$ such that the second term in equation \eqref{eq:lambda_neoW} disappears and invasion depends only on the sign of $(r-R)$. 
Indeed, invasion typically occurs when the neo-W is more closely linked to the selected locus than the ancestral sex-determining region (Figure \ref{fig:Combination_Centimorgans}).
Secondly, we can simplify cases where invasion occurs despite $R>r$ using the special case where $R=1/2$ and $r<1/2$. 
In table \ref{tab:specialcases} we give the conditions where invasion occurs where we further assume that haploid selection only occurs during one phase in one sex (e.g., during male meiosis only) and dominance coefficients are equal in the two sexes, $h^\female=h^\male$. 
Where there is no gametic competition and meiotic drive in one sex only, an unlinked neo-W can invade as long as the same allele is favoured in male and female diploid selection ($s^\female s^\male>0$, see Figure \ref{fig:Combination_Centimorgans}B), which is 50\% of the parameter space. 
Where there is no meiotic drive and gametic competition occurs in one sex only, an unlinked neo-W can invade as long as the same allele is favoured in male and female diploid selection and there are sex differences in selection of one type (e.g., $s^\female(s^\male-s^\female)>0$, see Figure \ref{fig:Combination_Centimorgans}C,D), which is 25\% of the parameter space. 
These special cases indicate that neo-W invasion can occur for a relatively large fraction of parameter space, even if the neo-W is less tightly linked to the selected locus, $R>r$.

@@

Under Fisherian sex ratio selection, autosomal modifiers favour equal investment in male and female offspring, i.e., a 1:1 sex ratio \citep{Fisher:1930wy,Charnov:1982wg,West:2009we}. 
Therefore, a novel environmental sex-determiner that causes half of its carriers to become female and half to become male ($k=1/2$) will be in males half of the time and in females half of the time (like an autosome), and so should be favoured by sex ratio selection.
%As with an autosomal locus, these sex-determiners will be found in males half of the time and in females half of the time.
However, we find that the growth rate of a rare, dominant offspring-controlled neo-ESD allele that produces males or females with equal probability ($k=1/2$) is
\begin{equation}
\lambda_{ESD',XY} =1+ \frac{1}{2}\frac{(\lambda_{Y',XY}-1) + (\lambda_{W',XY}-1)}{2} \Big|_{R=1/2} + O\left(\epsilon^3\right),
%V_A \frac{r}{1-2r}  \frac{s_{f} \left( s^\male - 3 s^\female\right) }{8 \left( s_{f} + s_{m} \right)^2} t_{m}^2.
\label{eq:lambda_ESD}
\end{equation}
\noindent
We note that recombination between the selected locus and the novel sex-determining locus, $R$, doesn't enter into the $k=1/2$ results because sex is essentially randomized each generation, %as if the novel sex-determiner were on an unlinked site that half the time acts as a neo-$W$ and half the time as a neo-$Y$.
preventing associations from building up between allele $A$ and sex.

Equation \eqref{eq:lambda_ESD} shows that invasion by a novel `perfect' ESD (equal sex ratio, $k=1/2$) mutation is the same for an ancestrally XY or ZW system (since $\lambda_{Y',XY} = \lambda_{W',ZW}$, $\lambda_{W',XY} = \lambda_{Y',ZW}$).
Thus, by the same argument as above (if drive occurs in males only then the sex ratio is only biased when the ancestral sex-determination system is XY), Fisherian sex ratio selection does not explain invasion by an offspring-controlled neo-ESD locus. 
Rather, the neo-ESD gets half of the fitness of a feminizing mutation (neo-$W$) and half of the fitness of a masculizing mutation (neo-$Y$), but only has an effect one half of the time (the other half of the time it produces the same sex as the ancestral system would have, to leading order). 
The net result can be that perfect ESD will not invade, even if sex ratios are biased. 
For example, if there is haploid selection in males (either drive or pollen/sperm competition) but the conditions in table \ref{tab:specialcases} are not met, perfect ESD will not invade, even though it would equalize the zygotic sex ratio from an initially biased case (assuming $r<1/2$).

@@

Fisherian sex ratio selection is sometimes considered in terms of balancing parental investment in male versus female offspring \citep{Charnov:1982wg}.
In addition, under environmental sex-determination, the proportion of males/females is sometimes controlled by the mother, e.g., the proportion of eggs laid in warm versus cold environments could determine the sex ratio of offspring. 
We therefore also considered the invasion of a neo-sex-determining allele ($m$) in a model in which mothers that have at least one $m$ allele produce daughters with probability $k$. 
As with offspring-controlled ESD, for all $k\in\{0,1/2,1\}$, we find that invasion into an ancestrally XY system is the same as invasion into an ancestrally ZW system (at least up to order $\epsilon^3$), implying transitions between genetic sex determination and maternally controlled environmental sex determination are not driven by Fisherian sex ratio selection on biased zygotic sex ratios.

@@

Taken at face value, our results indicate that transitions in heterogametey (XY to ZW or vice versa) are more likely to be favoured by selection if there is selection upon both haploid and diploid genotypes rather than diploid selection alone. 
This prediction could be examined using a suitable proxy for haploid selection, for example, \cite{Lenormand:2005vb} use the outcrossing rate as a proxy for the strength of pollen competition. 
In animals, one might expect gametic competition to be stronger in species where sperm is required to live for a long time after spermatogenesis because transcripts shared during spermatogenesis may become depleted, revealing the haploid phenotype of the sperm \citep{Immler:2014im}. 
Given the caveats mentioned above about the form of meiotic drive modelled, we would also expect that heterogametic transitions in sex determination would be more common in clades where there is meiotic drive.

@@

Selection among diploids then occurs according to the diploid genotype at the \textbf{A} locus, $h$, for an individual of type $ij$ ($h \in AA$, $Aa$, $aa$, see Table \ref{tab:fitnesstable}). 
The diploid frequencies after selection in sex $d$ are given by $xx_{ij}^{\Hermaphrodite,s}=w_{h}^{\Hermaphrodite} xx_{ij}/\bar{w}^{\Hermaphrodite}$, $xy_{ij}^{\Hermaphrodite,s}=w_{h}^{\Hermaphrodite} xy_{ij}/\bar{w}^{\Hermaphrodite}$, and $yy_{ij}^{\Hermaphrodite,s}=w_{h}^{\Hermaphrodite} yy_{ij}/\bar{w}^{\Hermaphrodite}$, where $\bar{w}^{\Hermaphrodite}= \sum_{i=1}^{4}\sum_{j=1}^{4}w_{h}^{\Hermaphrodite}xx_{ij}+w_{h}^{\Hermaphrodite}xy_{ij}+w_{h}^{\Hermaphrodite}yy_{ij}$ is the mean fitness of individuals of sex $\Hermaphrodite$.
